---
author: Nick Tustison
title: An open-source framework for multi-modal pulmonary image analysis
shorttitle:  Lung imaging and ANTs
bibliography: ../references.bib
logo:
output:
  beamer_presentation:
    colortheme: dolphin
    fonttheme:  structurebold
    highlight: tango
    incremental: no
    theme:  AnnArbor
    template: default2.tex
    includes:
      in_header: header.tex
    toc: no
    keep_tex: yes
  pdf_document:
    toc: true
    highlight: zenburn
  html_document:
    toc: true
    theme: readable
  revealjs_presentation:
    pandoc_args: [ "--slide-level", "2" ]
    incremental: true
    widescreen: true
    smaller: false
    theme: night
    transition: fade
    highlight: zenburn
    center: true
    self_contained: false
  ioslides_presentation:
    css: atk.css
    incremental: false
    widescreen: true
    smaller: false
    fig_width: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( cache=TRUE )
```

# Motivation

## What does the neuroimaging community offer?

Great packages such as:

* AFNI
* FSL
* FreeSurfer
* SPM
* ANTs

## Public & robust software $\longrightarrow$ research output

```{r pubmedQuery, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, results="hide", fig.height=3, fig.width=6}
years <- 2005:2016

numberOfAbstractsPerYear <- rep( 0, length( years ) )
for( i in 1:length( years ) )
  {
  cat( "Doing year", years[i], "\n" )
  query = "afni OR fsl OR freesurfer OR (\"statistical parametric mapping\") OR (\"advanced normalization tools\")"
  res <- EUtilsSummary( query, db = "pubmed", retmax = 5000, mindate = as.character( years[i] ), maxdate = as.character( years[i] ) )
  summary( res )
  fetch <- EUtilsGet( res )
  abstracts <- AbstractText( fetch )
  numberOfAbstractsPerYear[i] <- length( abstracts )
  }

plotDataFrame <- data.frame( Year = as.factor( years ), NumberOfAbstracts = numberOfAbstractsPerYear )
ggplot( data = plotDataFrame, aes( x = Year, y = NumberOfAbstracts, fill = NumberOfAbstracts ) ) +
          geom_bar( stat = "identity" ) +
          scale_fill_gradient( low = "#01256e", high = "#6d0e0e" ) +
          ylab( "Number of articles" ) +
          xlab( "Publication year" ) +
          theme( legend.position = "none" )
```

## Benefits of open-source:

* Motivates community-based support:
     * bug fixes (_"Given enough eyeballs, all bugs are shallow."_),
     * new features,
     * reproducibility verification, and
     * community tech support.
* Learn directly from journal manuscripts *and* implementations.
* Tremendous cost-savings.

##  E. A. Hoffman et al., JMRI 2015.

> "More widespread use of all [pulmonary] imaging biomarkers has been limited for a number
> of key reasons, including:
>        1) lack of support to harmonize image acquisition software;
>      __2) universally available image analysis software;__
>        3) regulatory boundaries for emerging approaches; and
>        4) historically weak links between respiratory and radiology clinical programs."

# ITK-Lung: A software framework for lung image processing and analysis

## ANTs core tools for lung image analysis

\centering
\begin{figure}
\includegraphics[width=0.95\textwidth]{./lung/figures/coreANtsToolsLung.png}
\end{figure}

## Proposed core functionality

|  __Functionality__  |  __CT__   |   __1H MRI__   | __3He MRI__ | __PET__
|:----|:-------:|:------:|:------:|:------:|
|   registration     |  $\circ$  |  $\circ$  |  $\circ$  |  $\circ$  |
|   template generation    |  $\circ$  |  $\circ$  |  $\circ$  |  $\circ$  |
|   lung segmentation    |  $\circ$  |  $\circ$  |  $\ddagger$  |  $\ddagger$  |
|   lobe segmentation    |  $\circ$  |  $\circ$  |  $\ddagger$  |  $\ddagger$  |
|   airway segmentation   |  $\circ$  |  --  |  --  |  --  |
|   vessel segmentation   |  $\circ$  |  --  |  --  |  --  |
|   functional segmentation   |  $\ast$  |  --  |  $\circ$  |  $\ast$  |
|   feature indices   |  $\circ$  |  --  |  $\ast$  |  $\ast$  |

\footnotesize

'$\circ$': previously published work

'$\ast$': cross-modality functionality

'$\ddagger$': simultaneous structural acquisitions



## EMPIRE 2010

\centering
\begin{figure}
  \includegraphics[width=0.9\textwidth]{./competitions/figures/EMPIRE10criteria.png}
\end{figure}

\centering
\begin{figure}
  \includegraphics[width=0.95\textwidth]{./competitions/figures/EMPIRE10Results.png}
\end{figure}

## Registration github examples

* https://github.com/ntustison/antsCtLungRegistrationExample

* https://github.com/ntustison/ProtonCtLungMaskRegistration

## Atlas-based lung and lobe estimation

\centering
\begin{figure}
\includegraphics[width=0.95\textwidth]{./lung/figures/malfLungsBoth.pdf}
\end{figure}

## Good results on 1H MRI

|     |  __Mean__   |   __SD__   | __Median__ |
|:----|:-------:|:------:|:------:|
| __Left lung__   |  0.963 (0.974)  |  0.013 (0.097)  |  0.964 (0.992) |
| __Right lung__  |  0.968 (0.972)  |  0.012 (0.135)  |  0.970 (0.996) |
| __Left upper__  |  0.882 (0.922)  |  0.059 (0.163)  |  0.894 (0.978) |
| __Left lower__  |  0.868 (0.885)  |  0.06 (0.229)   |  0.892 (0.964) |
| __Right upper__ |  0.852 (0.921)  |  0.067 (0.088)  |  0.875 (0.96)  |
| __Right middle__|  0.657 (0.765)  |  0.130 (0.299)  |  0.696 (0.886) |
| __Right lower__ |  0.873 (0.914)  |  0.063 (0.176)  |  0.900 (0.968) |

(*) Comparison with state-of-the-art

## Lung and lobe estimation github example

https://github.com/ntustison/LungAndLobeEstimationExample

## Functional ventilation

\centering
\begin{figure}
\includegraphics[width=0.9\textwidth]{./lung/figures/prePostAlbuterol.png}
\end{figure}

## Functional segmentation github example

https://github.com/ntustison/LungVentilationSegmentationExample

## Combining ANTs lung functionality

\centering
\begin{figure}
\includegraphics[width=0.9\textwidth]{./lung/figures/longitudinalStudy.png}
\end{figure}


<!--
## Combining ANTs lung functionality II

\centering
\begin{figure}
\includegraphics[width=0.9\textwidth]{./lung/figures/airways.png}
\end{figure}
-->
