---
author: |
   | Jim Gee and Nick Tustison
   | \vspace{0.25em} UPenn/UVa
title: |
   | An open-source framework for multi-modal pulmonary image analysis
   | \vspace{0.5em} \footnotesize  _ITK-Lung: A Software Framework for Lung Image Processing and Analysis_
   | \footnotesize (R01 HL133889-01A1)
   | \normalsize \vspace{3em} The 2017 International Workshop on Pulmonary Imaging
shorttitle:  Pulmonary open-science imaging
bibliography: ../references.bib
logo:
output:
  beamer_presentation:
    colortheme: dolphin
    fonttheme:  structurebold
    highlight: tango
    incremental: no
    theme:  AnnArbor
    template: default2.tex
    includes:
      in_header: header.tex
    toc: no
    keep_tex: yes
  pdf_document:
    toc: true
    highlight: zenburn
  html_document:
    toc: true
    theme: readable
  revealjs_presentation:
    pandoc_args: [ "--slide-level", "2" ]
    incremental: true
    widescreen: true
    smaller: false
    theme: night
    transition: fade
    highlight: zenburn
    center: true
    self_contained: false
  ioslides_presentation:
    css: atk.css
    incremental: false
    widescreen: true
    smaller: false
    fig_width: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( cache=TRUE )
```

# Why should you care?

## List of collaborators

|  |  |
|:----|:-------:|
| Mike Shim  | Grace Parraga  |
| Gerry Teague | Edwin van Beek |
| Tally Altes | Yoshiharu Ohno |
| Rahim Rizi | Joon-Beom Seo |
| Eduardo Barbosa |  Hans-Ulrich Kauczor |
| Warren Gefter | Jim Wild  |
| David Mankoff | Mark Scheibler |
| Sean Fain | Eric Hoffman |

##  

> "More widespread use of all [pulmonary] imaging biomarkers has been limited for a number
> of key reasons, including:
>        1) lack of support to harmonize image acquisition software;
>      __2) universally available image analysis software;__
>        3) regulatory boundaries for emerging approaches; and
>        4) historically weak links between respiratory and radiology clinical programs."

\begin{flushright}
--- E. A. Hoffman et al., JMRI 2015.
\end{flushright}

## "publications $=$ advertisements"

> "An article about computational science in a scientific publication is not the
> scholarship itself, it is merely advertising of the scholarship. The actual
> scholarship is the complete software development environment and the complete
> set of instructions which generated the figures."

\begin{flushright}
 --- Jonathan Buckheit and David Donoho (Jon Claerbout)
\end{flushright}

## Competitions

\centering
\begin{figure}
\includegraphics[width=0.95\textwidth]{./Figs/competitions.png}
\end{figure}


## What does the neuroimaging community offer?

Great packages such as:

* AFNI
* FSL
* FreeSurfer  $\longrightarrow$  NeuroQuant\textsuperscript{\textregistered}
* SPM
* ANTs

## Public & robust software $\longrightarrow$ research output

```{r pubmedQuery, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, results="hide", fig.height=3, fig.width=6}
years <- 2005:2016

numberOfAbstractsPerYear <- rep( 0, length( years ) )
for( i in 1:length( years ) )
  {
  cat( "Doing year", years[i], "\n" )
  query = "afni OR fsl OR freesurfer OR (\"statistical parametric mapping\") OR (\"advanced normalization tools\")"
  res <- EUtilsSummary( query, db = "pubmed", retmax = 5000, mindate = as.character( years[i] ), maxdate = as.character( years[i] ) )
  summary( res )
  fetch <- EUtilsGet( res )
  abstracts <- AbstractText( fetch )
  numberOfAbstractsPerYear[i] <- length( abstracts )
  }

plotDataFrame <- data.frame( Year = as.factor( years ), NumberOfAbstracts = numberOfAbstractsPerYear )
ggplot( data = plotDataFrame, aes( x = Year, y = NumberOfAbstracts, fill = NumberOfAbstracts ) ) +
          geom_bar( stat = "identity" ) +
          scale_fill_gradient( low = "#01256e", high = "#6d0e0e" ) +
          ylab( "Number of articles" ) +
          xlab( "Publication year" ) +
          theme( legend.position = "none" )
```

## Benefits of open-source:

* Motivates community-based support:
     * bug fixes (_"Given enough eyeballs, all bugs are shallow."_),
     * new features,
     * reproducibility, and
     * community tech support.
* Learn directly from journal manuscripts *and* implementations.
* Tremendous cost-savings.
* _"Don't reinvent the wheel."_

## ANTs core tools for neuroimage analysis

\centering
\begin{figure}
\includegraphics[width=0.95\textwidth]{../Figs/coreANtsToolsNeuro.png}
\end{figure}

# ITK-Lung

## ANTs core tools for lung image analysis

\centering
\begin{figure}
\includegraphics[width=0.95\textwidth]{./lung/figures/coreANtsToolsLung.png}
\end{figure}

## Proposed core functionality

|  __Functionality__  |  __CT__   |   __1H MRI__   | __3He MRI__ | __PET__ |
|:----|:-------:|:------:|:------:|:------:|
|   registration     |  $\circ$  |  $\circ$  |  $\circ$  |  $\circ$  |
|   template generation    |  $\circ$  |  $\circ$  |  $\circ$  |  $\circ$  |
|   lung segmentation    |  $\circ$  |  $\circ$  |  $\ddagger$  |  $\ddagger$  |
|   lobe segmentation    |  $\circ$  |  $\circ$  |  $\ddagger$  |  $\ddagger$  |
|   airway segmentation   |  $\circ$  |  --  |  --  |  --  |
|   vessel segmentation   |  $\circ$  |  --  |  --  |  --  |
|   functional segmentation   |  $\ast$  |  --  |  $\circ$  |  $\ast$  |
|   feature indices   |  $\circ$  |  --  |  $\ast$  |  $\ast$  |

\footnotesize

'$\circ$': previously published work

'$\ast$': cross-modality functionality

'$\ddagger$': simultaneous structural acquisitions
